Automatic Identification and Annotation of Tasks in Structured Programs

### 2018-04-10 docs/ACM/paper.tex
Fixing the paper in 11 pages.

\begin{abstract}
This paper reports the design and implementation of a suit of static analyses
and code generation techniques to annotate programs with OpenMP pragmas for
task parallelism.
These techniques approximate the ranges covered by memory regions, bound
recursive tasks and estimate the profitability of tasks.
We have used these ideas to implement a source-to-source compiler that inserts 
OpenMP pragmas into C/C++ programs without any human intervention.
By building onto the static program analysis literature, and relying on OpenMP's
runtime ability to disambiguate pointers, we show that we can annotate large and
convoluted programs, often replicating the performance gains of handmade
annotation.
Furthermore, our techniques give us the means to discover, without any 
development cost, opportunities of parallelism that remained buried in the syntax 
of well-known benchmarks for many years.
\end{abstract}

@@

This paper describes \Taskminer, a source-to-source compiler that exposes task
parallelism in C/C++ programs.
To fulfil this goal, \Taskminer{} solves different challenges.
First, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:identification}).
Second, it determines symbolic bounds to the memory blocks accessed within
those regions (Section~\ref{sub:sra}).
Third, it extracts parameters from the code to estimate when it is profitable to
create tasks.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}) and limit their recursion
depth (Section~\ref{sub:ir}).
Fourth, \Taskminer{} determines which program variables need to be 
privatized in new tasks, or shared among them (Section~\ref{sub:variance}).
Finally, it maps all this information back into source code, producing
readable annotations (Section~\ref{sub:ir}).

In this paper, we defend the thesis that automatic task annotations are
effective and useful.
Our techniques enhance the productivity of developers, because they save them
the time to annotate programs.
\Taskminer{} receives as input C code, and produces, as output, a C
program annotated with human-readable OpenMP task directives.
We are currently able to annotate non-trivial programs, involving every
sort of composite type available in the C language, e.g., arrays, structs,
unions and pointers to these aggregates.
Some of these programs, such as those taken from the
Kastor~\cite{Virouleau14} or Bots~\cite{Duran09}
benchmark suites, are large and complex; thus, their manual annotation is a
time consuming and error prone task.
Yet, our automatically annotated programs not only approximate the execution
times of the parallel versions of those benchmarks, but are much faster than
\guido{Numbers?}
their sequential --unannotated-- versions, as we report in Section~\ref{sec:eval}.

@@

Most of the benchmarks, e.g., 27, that have been automatically annotated did not
give us speedups, because the regions marked as tasks were too small to influence
the program's runtime.
We have observed slowdowns in 17 benchmarks.
In this case, interactions between data-structures end up forcing the OpenMP
runtime to serialize execution of tasks.
Most of these slowdowns were inferior to 10\%.
In one case, \textsf{MiBench/office-stringsearch}, our program was 11x slower.
Nevertheless, we have measured speedups above 5\% in 19 benchmarks.
In \textsf{Misc/lowercase}, the speedup is above 3.5x, and in
three cases, it is above 1.5x.
Figure~\ref{fig:TM_GeneralProgs} summarizes these findings.
We emphasize that this experiment did not involve any human intervention.
Hence, {\em \Taskminer{} enables the discovery of parallelism at
zero programming cost}.

@@

The analyses that we have described in Section~\ref{sec:sol} have a worst-case
quadratic time.
This is the worst-case scenario of the symbolic range analysis of
Section~\ref{sub:sra} and the task-expansion algorithm of
Section~\ref{sub:expansion}.
Nevertheless, our implementation is fast in practice.
To support this statement, we have used \textsf{CSmith}~\cite{Yang11}, a random
code generator, to produce 100 programs of varying sizes, which we then fed to
the \Taskminer.
Figure~\ref{fig:TM_Runtime} shows the result of this experiment.
Our largest program had 4,999 lines of C, and \Taskminer{} could process it
in 1.2 seconds.
the $R^2$ of polynomials of degree 1, 2, 3 and 4 is, respectively, 0.795,
0.890, 0.909 and 0.911, suggesting very small differences between polynomials
of degree 2 or more.

@@

Mainstream compilers have only recently added support to OpenMP 4.0's task
parallelism.
An implementation of \textsf{clang} supporting tasks was released in March 14th,
2014\footnote{\url{https://clang-omp.github.io/}}.
A few weeks later, in April 22nd, 2014, such support was also announced in
\textsf{gcc} 4.9.
Thus, because the necessary infra-structure for the implementation of tasks is
relatively new, currently there are no tools, other than the \Taskminer,
that annotate programs automatically with these directives.
Nevertheless, there exists a long string of research aiming at the automatic
parallelization of programs.
In this section, we examine some elements in this list that relate to our
work.

@@

\pedro{As future work, we plan on performing Autotuning on {\Taskminer} to find good default parameters.
Autotuning works for data-flow applications have been already done in \cite{Trancoso17, Emani15}}.

>>>>

\begin{abstract}
This paper describes the design and implementation of a suit of static analyses
and code generation techniques to annotate programs with OpenMP pragmas for
task parallelism.
These techniques approximate the ranges covered by memory regions, bound
recursive tasks and estimate the profitability of tasks.
We have used these ideas to implement a source-to-source compiler that inserts 
OpenMP pragmas into C/C++ programs without any human intervention.
By building onto the static program analysis literature, and relying on OpenMP's
runtime ability to disambiguate pointers, we show that we can annotate large and
convoluted programs, often replicating the performance gains of handmade
annotation.
Furthermore, our techniques give us the means to discover, without any 
development cost, opportunities of parallelism that remained buried in the syntax 
of well-known benchmarks for many years.
\end{abstract}

@@

This paper describes \Taskminer, a source-to-source compiler that exposes task
parallelism in C/C++ programs.
To fulfil this goal, \Taskminer{} solves different challenges.
First, it determines symbolic bounds to the memory blocks accessed within
programs (Section~\ref{sub:sra}).
Second, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:identification}).
Third, it extracts parameters from the code to estimate when it is profitable to
create tasks.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}) and limit their recursion
depth (Section~\ref{sub:ir}).
Fourth, \Taskminer{} determines which program variables need to be 
privatized in new tasks, or shared among them (Section~\ref{sub:variance}).
Finally, it maps all this information back into source code, producing
readable annotations (Section~\ref{sub:ir}).

In this paper, we defend the thesis that automatic task annotations are
effective and useful.
Our techniques enhance the productivity of developers, because they save them
the time to annotate programs.
In Section~\ref{sec:eval}, we show that we have been able to achieve almost
four-fold speedups onto standard benchmarks (not-used in parallel programming),
with zero programming cost.
\Taskminer{} receives as input C code, and produces, as output, a C
program annotated with human-readable OpenMP task directives.
We are currently able to annotate non-trivial programs, involving every
sort of composite type available in the C language, e.g., arrays, structs,
unions and pointers to these aggregates.
Some of these programs, such as those taken from the
Kastor~\cite{Virouleau14}, Bots~\cite{Duran09} or the LLVM~\cite{Lattner04} test
suites, are large and complex.
Yet, our automatically annotated programs not only approximate the execution
times of the parallel versions of those benchmarks, but are much faster than
their sequential --unannotated-- versions, as we report in
Section~\ref{sec:eval}.

@@

Most of the benchmarks, e.g., 27, that have been automatically annotated did not
give us speedups, because the regions marked as tasks were too small to influence
the program's runtime.
We have observed slowdowns in 17 benchmarks.
In this case, interactions between data-structures end up forcing the OpenMP
runtime to serialize execution of tasks.
Most of these slowdowns were inferior to 10\%.
In one case, \textsf{MiBench/office-stringsearch}, our program was 11x slower.
Nevertheless, we have measured speedups above 5\% in 19 benchmarks.
In \textsf{Misc/lowercase}, the speedup is above 3.5x, and in
three cases, it is above 1.5x.
Figure~\ref{fig:TM_GeneralProgs} shows our 10 largest speedups.
We emphasize that this experiment did not involve any human intervention.
Hence, {\em \Taskminer{} enables the discovery of parallelism at
zero programming cost}.

@@

The analyses described in Section~\ref{sec:sol} have a worst-case quadratic time.
This is the worst-case scenario of the symbolic range analysis of
Section~\ref{sub:sra} and the task-expansion algorithm of
Section~\ref{sub:expansion}.
Nevertheless, our implementation is fast in practice.
To support this statement, we have used \textsf{CSmith}~\cite{Yang11}, a random
code generator, to produce 100 programs of varying sizes, which we then fed to
the \Taskminer.
Figure~\ref{fig:TM_Runtime} shows the result of this experiment.
Our largest program had 4,999 lines of C, and \Taskminer{} could process it
in 1.2 seconds.
the $R^2$ of polynomials of degree 1, 2, 3 and 4 is, respectively, 0.795,
0.890, 0.909 and 0.911, suggesting very small differences between polynomials
of degree 2 or more.

@@

Mainstream compilers have only recently added support to OpenMP 4.0's task
parallelism.
An implementation of \textsf{clang} supporting tasks was released in March 14th,
2014.
A few weeks later, in April 22nd, 2014, such support was also announced in
\textsf{gcc} 4.9.
Thus, because the necessary infra-structure for the implementation of tasks is
relatively new, currently there are no tools, other than the \Taskminer,
that annotate programs automatically with these directives.
Nevertheless, there exists a long string of research aiming at the automatic
parallelization of programs.
In this section, we examine some elements in this list that relate to our
work.

@@

This paper has described a methodology to annotate programs with task
parallel pragmas, which we could demonstrate to be effective
on general programs.
This methodology does not introduce any fundamentally new static analysis or code 
optimization; in this field, we claim no contribution.
Instead, our contributions lay into the overall design that eventually
emerged from a two-years long effort to combine existing compilation techniques
towards the goal of profiting from the powerful runtime system that OpenMP's
task parallelism brings forward.
During this period, we went through many dead-ends; too many, indeed, for a
11-pages report.
Nevertheless, this experience lets us single out a few elements from the
compiler literature, such as symbolic range analysis and windmills, which we
could use to solve challenges related to automatic code annotation.
Much work is still left to be done, until we can reach a stage in which
automatic annotations can beat consistently manual interventions.
In particular, our methodology asks for more aggressive tuning strategies.
Techniques such as Trancoso's~\cite{Trancoso17}, Iwasaki's~\cite{Iwasaki16B}
or Emani's~\cite{Emani15} have already been shown to be effective in this
domain, and we want to explore them further.
More specifically, a future release of \Taskminer{} is already expected to
feature Iwasaki's cut-off~\cite{Iwasaki16B}.
